---
title: "Regression Models - Course Project"
author: "adrianvs"
output: pdf_document
---

```{r, warning=FALSE, message=FALSE, echo=FALSE}
require(datasets); require(dplyr); require(ggplot2); require(car); require(xtable)
require(RColorBrewer); require(gridExtra)
options(xtable.comment = FALSE)
```

## Executive Summary
This report tries to answer two questions: which transmission type is better for MPG and
how strong is its imapact on fuel efficiency.
Although there is a significant difference in MPG between transmission types in the data set,
with manual transition beeing more efficient, once the data is adjusted for other
variables no significant difference remains. The estimated impact lies between -2.61 and 2.86
MPG for manual transmission (95% confidence interval for the manual transmission 
regression coefficient). The true impact can not be quantified from this data set.

## Exploratory Analysis

The mtcars data set describes 10 variables of 32 cars (1973 - 1974 models) published in 
the 1974 Motor Trend US magazine.
The observed varaibles are: miles per gallon, number of Cylinders, displacement, horsepower
, rear axle ratio, weight, 1/4 mile time, v-shaped or straight engine, transmisison type, 
number of gears and number of carburetors.
The following table gives a summary of the mileage with respect to transmission type.
```{r, echo=FALSE}
data(mtcars)
df <- mtcars
df$am <- factor(df$am, levels = c(0,1), labels = c("automatic", "manual"))
df$vs <- factor(df$vs)
df$carb <- factor(df$carb)

table1 <-
df %>% group_by(am) %>%
       summarise("n" = length(mpg), 
                 "Minimum" = min(mpg),  
                 "1st Quantile" = quantile(mpg,0.25),
                 "Median" = median(mpg),
                 "Mean" = mean(mpg),
                 "3rd Quantile" = quantile(mpg,0.75),
                 "Maximum" = max(mpg),
                 "Standard Deviation" = sd(mpg))
colnames(table1)[1] <- "Transmission"


```
```{r, echo=FALSE, results="asis"}
print(xtable(table1,align="rrcccccccc"), include.rownames=FALSE, floating = FALSE)  
```
<br><br>
*Table 1: Summary: Miles per gallon by Transmission type*  

**Figure 1** (appendix) shows, that if no other variables are accounted for there seems to be a distinct
advantage for manual transmission in miles per gallon.   


A closer look at the distribution
of other variables that have an impact on miles per gallon with respect
to transmission type point to a possible bias. Manual transmission cars in this data set 
seem to weigh less, have lower engine displacement, lower horsepower and lower acceleration 
(measured by quarter mile time). See **figure 2** of the appendix.
The unpaired, two sample t-test results are shown in table 2.

```{r, results='asis', echo = TRUE}
t1 <- t.test(df$mpg  ~ df$am); t2 <- t.test(df$wt   ~ df$am); t3 <- t.test(df$hp   ~ df$am)
t4 <- t.test(df$disp ~ df$am); t5 <- t.test(df$qsec ~ df$am)
```
```{r, results='asis', echo = FALSE}
table2 <- data.frame(rbind(
        Mpg = c(t1$conf.int[1], t1$conf.int[2], t1$p.value),
        Weight =c(t2$conf.int[1], t2$conf.int[2], t2$p.value),
        Horsepower = c(t3$conf.int[1], t3$conf.int[2], t3$p.value),
        Displacement = c(t4$conf.int[1], t4$conf.int[2], t4$p.value),
        QuarterMile = c(t5$conf.int[1], t5$conf.int[2], t5$p.value)))
names(table2) <- c("CI lower", "CI upper", "p-value")
print(xtable(table2,align="rlll", digits = c(2,2,2,3)), include.rownames=TRUE, floating = FALSE)  
```
<br><br>
*Table 2: T-Test: Manual vs Automatic Transmission*  

## Model Selection
```{r, echo = FALSE}
fitall <- lm(mpg ~ ., data = df)
x <- data.frame(summary(fitall)$coef) %>% summarise(n = n()-1, 
                                               biggestP = max(Pr...t..), 
                                               smallestP = min(Pr...t..))

```
To quantify the effect the mode of transmission has on fuel efficiency and account in part for
the observed bias in other variables correlated with it, we fit a linear model that has to includes
transmission type as a predictor.
Since this model is not intended for prediction, all quadratic and interaction terms including 
the transmission variable will be omitted to allow for easy interpretability of this coefficient. 
In a model using all varibles as regressors none remains a significant predictor of fuel efficiency,
displaying the effect of collinearity between the regressors.
Coefficients p-values lie between `r round(x$smallestP,2)` and `r round(x$biggestP,2)`
in this model.

```{r, results='asis', echo=FALSE}
a <- data.frame(with(mtcars, cor(am, mtcars[names(mtcars[-9])])))
names(a) <- c("Mpg","Cyl.", "Displ.", "HP", "Rear axle ratio", "Weight", 
              "1/4 Mile time", "Engine layout", "Gears", "Carburetors")

print(xtable(a[1,2:10],align="cccccccccc"), include.rownames=FALSE, floating = FALSE)  
 
```
<br><br>
*Table 3: Correlation of Varibles to Transmission Type*  


To adjust the effect of transmission type, variables, that are unevenly distributed between
manual and automatic transmission and have a high correlation to fuel efficiency, 
are included stepwise into the model.
```{r, results = 'asis'}
lm1 <- lm(mpg ~ am, data = df)
lm2 <- lm(mpg ~ am + wt, data = df)
lm3 <- lm(mpg ~ am + wt + hp, data = df)
lm4 <- lm(mpg ~ am + wt + hp + qsec, data = df)
lm5 <- lm(mpg ~ am + wt + hp + qsec + disp, data = df)
table3 <- anova(lm1,lm2,lm3,lm4,lm5)
```
```{r, results = 'asis', echo = FALSE}
print(xtable(table3,align="ccccccc"), include.rownames=FALSE, floating = FALSE)
```
<br><br>
*Table 4: Analysis of Variance - Nested Likelihood Ratio*   
 
As can be seen by the rise in p-values, the inclusion of quarter mile time and engine
disposition does not add significantly to the model but increases its variance.
The third model including weight and horsepower ist the strongest so far.
Since weight and horsepower are relatively strongly correlated, the inclusion of an 
interaction term in the linear model may better the fit.
```{r, results = 'asis'}
lmInt <- lm(mpg ~ am + wt + hp + wt:hp, data = df)
```
```{r, results = 'asis', echo=FALSE}

print(xtable(anova(lm3,lmInt),align="ccccccc"), include.rownames=FALSE, floating = FALSE)
```

*Table 5: Analysis of Variance(2) - Nested Likelihood Ratio*  

The inclusion of the interaction term adds significantly to the model. The residual sum of 
squares fell to below 130 and all coefficients are significant with pvalues from the F
statistic below 0.005 in the ANOVA table. The adjusted R-squared of the model is 0.8677. 
The **manual transmission coefficient is `r round(summary(lmInt)$coef[2,1],3)`** MPG more for manual
transmission cars with all other model regressors held constant and 
a **95% Confidence Interval of `r round(confint.lm(lmInt)[2,],2)`; no significant difference.**

## Residuals and Diagnostics
#### A. Residuals
The residual plot (**figure 3**, appendix) shows no obvious pattern in the residuals 
and confirms
that a linear model fits the data. The Quantile-Quantile plot demonstrates that the residuals
are nearly normally distributed.  

#### B. Homoscedasticity  
The Scale Location plot tests whether the variance of the residuals change as a function of the
fitted values. Since no clear trend in the red trend line is visible the residuals are 
assumed to be homoscedastic.  

#### C. Influential Outliers
No outlier has a Cook's Distance bigger than 0.2 and no high leverage, high standardized
residual is present (see figure 3 (bottom right), appendix).


## Appendix
<br><br>
```{r, fig.width=12, fig.height=7, echo=FALSE}

box <- 
 ggplot(df, aes(y = mpg, x = am, fill = am)) +
         geom_boxplot() + 
         #facet_grid(. ~ ) +
         scale_fill_brewer(palette="Set1") + 
         ggtitle(expression(atop("MPG by Transmission", 
                                 atop(italic("automatic vs manual"), "")))) +
         labs(x="Transmission", y="Milse per Gallon") +
         theme(plot.title = element_text(size = 28), axis.title = element_text(size = 20)) +
         guides(fill=FALSE) +
         stat_summary(fun.y=mean, geom="point", shape=5, size=2) +
         theme_classic(base_size = 14)
 
dens <- 
 ggplot(df, aes(x=mpg, fill = am)) +
         geom_density(colour="black", alpha = 0.7)  +
         scale_fill_brewer(palette="Set1") + 
         ggtitle(expression(atop("Density Curve", 
                                 atop(italic("automatic vs manual"), "")))) +
         labs(x="Transmission", y="Milse per Gallon") +
         theme(plot.title = element_text(size = 28), axis.title = element_text(size = 20)) +
         labs(x="Miles per Gallon", y="Count")  +
         guides(fill=FALSE) +
         theme_classic(base_size = 14)


grid.arrange(box,dens, ncol = 2)
```
<br><br>
*Figure 1: Miles per Gallon by Transmission Type*  
<br><br>
```{r,fig.width=12, fig.height=12, echo=FALSE}
p1 <-
ggplot(df, aes(y = hp, x = am, fill = am)) +
        geom_boxplot() + 
        scale_fill_brewer(palette="Set1") + 
        ggtitle(expression(atop("Horsepower by Transmission", 
                                atop(italic("automatic vs manual"), "")))) +
        labs(x="Transmission", y="Horsepower") +
        theme(plot.title = element_text(size = 28), axis.title = element_text(size = 20)) +
        guides(fill=FALSE) +
        stat_summary(fun.y=mean, geom="point", shape=5, size=2) +
        theme_classic(base_size = 14)

p2 <- 
ggplot(df, aes(y = wt/2, x = am, fill = am)) +
        geom_boxplot() + 
        scale_fill_brewer(palette="Set1") + 
        ggtitle(expression(atop("Weight by Transmission", 
                                atop(italic("automatic vs manual"), "")))) +
        labs(x="Transmission", y="Weight [US tons]") +
        theme(plot.title = element_text(size = 28), axis.title = element_text(size = 20)) +
        guides(fill=FALSE) +
        stat_summary(fun.y=mean, geom="point", shape=5, size=2) +
        theme_classic(base_size = 14)

p3 <- 
ggplot(df, aes(y = qsec, x = am, fill = am)) +
        geom_boxplot() + 
        scale_fill_brewer(palette="Set1") + 
        ggtitle(expression(atop("Quarter Mile Time by Transmission", 
                                atop(italic("automatic vs manual"), "")))) +
        labs(x="Transmission", y="Quarter Mile Time [s]") +
        theme(plot.title = element_text(size = 28), axis.title = element_text(size = 20)) +
        guides(fill=FALSE) +
        stat_summary(fun.y=mean, geom="point", shape=5, size=2) +
        theme_classic(base_size = 14)

p4 <- 
ggplot(df, aes(y = disp, x = am, fill = am)) +
        geom_boxplot() + 
        facet_grid(. ~ cyl) + 
        scale_fill_brewer(palette="Set1") + 
        ggtitle(expression(atop("Engine Displacement by Transmission and Cylinder", 
                                atop(italic("automatic vs manual"), "")))) +
        labs(x="Transmission", y="Engine Displacement [cubic inches]") +
        theme(plot.title = element_text(size = 28), axis.title = element_text(size = 20)) +
        guides(fill=FALSE) + 
        stat_summary(fun.y=mean, geom="point", shape=5, size=2) +
        theme_classic(base_size = 14)


grid.arrange(p1,p2,p3,p4, ncol = 2)


```
<br><br>
*Figure 2: Distribution of Variables by Transmission Type*  
<br><br>

```{r, fig.width=12, fig.height=12, echo = FALSE}

residplot <- 
  ggplot(lmInt, aes(x=.fitted, y=.resid)) +
        geom_point(size = 3) +
        geom_hline(yintercept = 0, col="black", linetype="dashed") +
        stat_smooth(method="loess", color = "red") +
        ggtitle(expression(atop("Residuals vs Fitted Values", 
                                atop(italic("Interaction Model"), "")))) +
        labs(x="Fitted Values", y="Residuals") +
        theme(plot.title = element_text(size = 28), axis.title = element_text(size = 20)) +
        theme_classic(base_size = 14)


qqres <-
  ggplot(lmInt, aes(qqnorm(.stdresid)[[1]], .stdresid))+
        geom_point(size = 3) +
        geom_abline(aes(qqline(.stdresid))) +
        ggtitle(expression(atop("QQ - Plot", 
                                atop(italic("Interaction Model"), "")))) +
        labs(x="Theoretical Quantiles", y="Standardized Residuals") +
        theme(plot.title = element_text(size = 28), axis.title = element_text(size = 20)) +
        theme_classic(base_size = 14)

reslev <-
  ggplot(lmInt, aes(.hat, .stdresid)) +
        geom_point(aes(size=.cooksd), na.rm=TRUE) +
        stat_smooth(method="loess", color = "red") +
        labs(x="Leverage", y="Standardized Residuals") +
        ggtitle(expression(atop("Residual vs Leverage Plot", 
                                atop(italic("Interaction Model"), "")))) +
        scale_size_continuous("Cook's Distance", range=c(1,5)) +
        theme(plot.title = element_text(size = 28), axis.title = element_text(size = 20)) +
        theme_classic(base_size = 14) +
        theme(legend.position="bottom")

scaleloc <-
  ggplot(lm3, aes(.fitted, sqrt(abs(.stdresid)))) +
        geom_point(size = 3) +
        stat_smooth(method="loess", color = "red") +
        labs(x="Fitted Value", y=expression(sqrt("|Standardized residuals|"))) +
        ggtitle(expression(atop("Scale Location", 
                                atop(italic("Interaction Model"), "")))) +
        theme(plot.title = element_text(size = 28), axis.title = element_text(size = 20)) +
        theme_classic(base_size = 14) 

grid.arrange(residplot,qqres,scaleloc,reslev, ncol = 2)
```
<br><br>
*Figure 3: Residuals and Diagnostic Plots*  
<br><br>  
<br><br>
The diagnostic plots used adapted ggplot2 code from **rpubs.com/therimalaya/43190** 
